Climate variability indices for ecological and crop models in R: the `climatrends` package

### 2020-10-02 inst/paper/paper.Rmd
paper-remove-ecological-analysis

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: false
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

# Introduction

Abiotic factors play an important role in most ecological and crop systems that depend on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understanding how these factors drive the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation.

[...continue...]

`climatrends` aims to provide the R [@RCoreTeam] toolkit to compute extreme precipitation and temperature indices that serve as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

[...continue...]

@@

Six main functions are provided, `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector' and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The last two methods are designed to fetch data from cloud sources.

[...continue...]

say that the idea for the functions started with citizen science projects that is why `day.one` and `span` may be variable across locations. For time series analysis where fixed periods are defined across many locations the indices can be adjusted with last.day.

[...continue...]

@@

The indices computed by the functions `temperature()` and `rainfall()` are described in Table 2.

@@

# Figures and tables

>>>>

---
title: 'Climate variability indices for ecological and crop models in R: the `climatrends` package'
tags:
- climate data
- climatology
- earth science
- evapotranspiration
- precipitation data
- R
- reproducibility
- weather data
authors:
  - name: Kauê de Sousa
    orcid: 0000-0002-7571-7845
    affiliation: "1, 2"
  - name: Magne Neby
    affiliation: "3"
  - name: Jacob van Etten
    orcid: 0000-0001-7554-2558
    affiliation: 2
  - name: Svein Ø. Solberg
    orcid: 0000-0002-4491-4483
    affiliation: 1
affiliations:
  - name: Department of Agricultural Sciences, Inland Norway University of Applied Sciences, Hamar, Norway
    index: 1
  - name: Bioversity International, Rome, Italy
    index: 2
  - name: Department of Forest and Field Sciences, Inland Norway University of Applied Sciences, 2480 Koppang, Norway
    index: 3
citation_author: de Sousa et. al.
date: "`r format(Sys.time(), '%d %B %Y')`"
year: 2020
bibliography: paper.bib
csl: citationstyle.csl
output:
  pdf_document:
    keep_tex: false
header-includes: \usepackage{caption} \captionsetup[figure]{labelformat=empty}
---

@@

# Summary

Abiotic factors play an important role in most ecological and crop systems that depend on certain levels of temperature, light and precipitation (and their interplay) to initiate important physiological events [@PlantEcology]. In the walk of climate change, understanding how these factors drive the physiological processes is a key approach to provide recommendations for adaptation and biodiversity conservation. The package `climatrends` aims to provide the methods in R [@RCoreTeam] to compute precipitation and temperature indices that serve as input for climate and crop models [@vanEtten2019; @Kehel2016], trends in climate change [@Aguilar2005; @deSousa2018] and applied ecology [@Prentice1992; @YLiu2018].

@@

Six main functions are provided (Table 1), `crop_sensitive()`, `ETo()`, `GDD()`, `late_frost()`, `rainfall()` and `temperature()` with a default method for numeric 'vector' and additional methods implemented via the package `methods` [@RCoreTeam] for classes 'matrix' (or array), 'data.frame', and 'sf' (of geometry POINT or POLYGON) [@sf]. The last two methods are designed to fetch data from cloud sources, until now from the packages `nasapower` [@Sparks2018] and `chirps` [@chirps].

These functions started as a set of scripts to compute indices in citizen science trials. In these trials. Aiming to capture the environmental variation across different sites, which can differ since each data-point generally have a different starting day and duration, the arguments `day.one` and `span` are vectorised and may be variable across data-points. For time series analysis, where fixed periods are defined across many locations, the indices can be adjusted with the argument `last.day` linked to the argument `day.one`.

@@

The package `climatrends` computes 12 temperature indices and 10 precipitation indices that were suggested by previous research on climatology and crop science [@Kehel2016; @Aguilar2005]. The indices computed by the functions `temperature()` and `rainfall()` are described in Table 2.

@@
